---
title: "exploratory_analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(here)
library(lubridate)
library(forecast)
library(MLmetrics)
library(tsibble)
library(astsa)
library(janitor)
library(feasts)
library(gt)
library(modelr)
library(astsa)
```



Load in Honolii Sea Surface Temperature Data using here package

```{r, results='hide'}
sst_honolii_1985_2021 <- read.csv(here("data", "sst_honolii_1985_2021.csv" ))
sst_honolii_1985_2021
```


Clean Data

```{r, results='hide'}
sst_honolii_1985_2021 <- sst_honolii_1985_2021 %>% 
  select(time, CRW_SST) %>% 
  mutate(time = as.Date(time))
sst_honolii_1985_2021
```

Determine SST Monthly Means

```{r, results='hide'}
sst_honolii_1985_2021_month <- sst_honolii_1985_2021 %>% 
  mutate(time = floor_date(time, "month")) %>% 
  group_by(time) %>% 
  summarise(mean_sst = mean(CRW_SST))

sst_honolii_1985_2021_month
```

Plot Data

```{r}
sst_honolii_1985_2021_month_plot <- ggplot(data = sst_honolii_1985_2021_month, aes(x = time, y = mean_sst)) +
  geom_line()

sst_honolii_1985_2021_month_plot
```

To estimate the linear relationship between time and mean sea surface temperature, we’ll use lm() to fit a regression model using Ordinary Least Squares (OLS).

```{r}
lm(mean_sst ~ time, data = sst_honolii_1985_2021_month) %>% 
  tidy %>% 
  gt()
```


Use the geom_smooth() function with method argument set to lm, which stands for “linear model”, to add the best fit OLS line to the sst_honolii_1985_2021_month_plot
```{r}
sst_honolii_1985_2021_month_regression <- sst_honolii_1985_2021_month_plot +
  geom_smooth(method="lm", col="black")

sst_honolii_1985_2021_month_regression
```
```{r}
ols <- lm(mean_sst ~ time, data = sst_honolii_1985_2021_month)
ols
summary(ols)

predictions <- sst_honolii_1985_2021_month %>% 
  add_predictions(ols) %>%
  mutate(residuals = mean_sst-pred)

# histogram
ggplot(data=predictions) + geom_histogram(aes(residuals), bins=25)

# mean
mean(predictions$residuals)
#1.495545e-14

# variance in residuals against measn_sst
ggplot(predictions) + geom_point(aes(x=mean_sst, y=residuals)) 

```




Convert dataframe into a time series
```{r, results='hide'}
sst_honolii_1985_2021_month_ts <- ts(sst_honolii_1985_2021_month$mean_sst, start = c(1985, 11), frequency = 12)

sst_honolii_1985_2021_month_ts
```


Convert time series into a tsibble to run more analysis

```{r}
sst_honolii_1985_2021_month_tsibble <- sst_honolii_1985_2021_month_ts %>% as_tsibble() 
sst_honolii_1985_2021_month_tsibble
```


Run a classical decomposition model

```{r}
sst_honolii_1985_2021_month_tsibble %>%
  model(
    classical_decomposition(value, type = "additive")
  ) %>%
  components() %>%
  autoplot() +
  labs(title = "Classical additive decomposition")
```


View the trend line on the data


```{r}
dcmp <- sst_honolii_1985_2021_month_tsibble %>%
  model(stl = STL(value))
components(dcmp)

components(dcmp) %>%
  as_tsibble() %>%
  autoplot(value, colour="gray") +
  geom_line(aes(y=trend), colour = "#D55E00")
```



An alternative plot that emphasizes the seasonal patterns is where the data for each season are collected together in separate mini time plots.

```{r}
sst_honolii_1985_2021_month_tsibble %>% 
  gg_subseries(value)
```


Forecasting

Set up training and validation time periods

```{r}
training=window(sst_honolii_1985_2021_month_ts, start = c(1985,11), end = c(2016,11))
validation=window(sst_honolii_1985_2021_month_ts, start = c(2016,11), end = c(2021,10))
```

DSHW Model from the forecast package

The DSHW Model returns forecasts using Taylor's (2003) Double-Seasonal Holt-Winters method.
  1. period1 = shorter seasonal period. 
    a. This period is shorter and occurs more often than period 2. It is often set to 4 if period 2 is set to 12 to represent four three month     long seasons in a 12 month year. (These periods do not have to be set to 4 and 12, one just needs to be an integer multiple of the other).
  2. period2 = longer seasonal period
  3. alpha, beta, gamma, omega = smoothing parameters (If NULL, the parameter is estimated using least squares.) 
  4. MAPE = mean average percentage error 


DSHW Model Fit
NOTE try to increase periods without messing up time
```{r}
dshw_model = dshw(training, period1=4, period2=12, h=length(validation))
MAPE(dshw_model$mean, validation)*100

summary(dshw_model)

dshw_plot <-plot(sst_honolii_1985_2021_month_ts, col="blue", xlab="Day", ylab="SST", main="DSHW Fit", type='l')
lines(dshw_model$mean, col="red", lwd=2)
```



DSHW Forecast

```{r}
fcst_dshw <- forecast(dshw_model_forecast, h = 120)
autoplot(fcst_dshw, include = 220)
print(summary(fcst_dshw))

```




ARIMA/SARIMA

ARIMA models contain three things:
AR (p): autoregressive part of the model. Means that we use p past observations from the timeseries as predictors.
Differencing (d): Used to transform the timeseries into a stationary one by taking the differences between successive observations at appropriate lags d.
MA (q): uses q past forecast errors as predictors.

SARIMA adds a seasonal part to the model to give us (p,d,q) (P,D,Q)

P = Seasonal Autoregressive Order(p)
D = Seasonal Differencing Order(d)
Q = Seasonal MA Order(q) 
m = Number of observations per year

auto.arima() - tries many different arima methods and returns the model that fits the best regarding (p,d,q) (P,D,Q)m
   Example: Best model: ARIMA(1,1,1)(2,1,1)[12]


```{r}
fit_arima_training <- auto.arima(training, d = 1, D = 1, stepwise = FALSE, approximation = FALSE, trace = TRUE)
print(summary(fit_arima_training))
checkresiduals(fit_arima_training)

test_forecast <- forecast(fit_arima_training, h = 120)
autoplot(test_forecast, include = 120)
```


```{r}
sarima_forecast = sarima.for(training, n.ahead=length(validation),
                              p=1,d=1,q=1,P=2,D=1,Q=1,S=12)
MAPE(sarima_forecast$pred, validation) * 100

sarima_plot <-plot(sst_honolii_1985_2021_month_ts, col="blue", xlab="Day", ylab="SST", main="SARIMA Fit", type='l')
lines(sarima_forecast$pred, col="red", lwd=2)
```

```{r}
fcst_arima <- forecast(fit_arima_training, h = 146, find.frequency = TRUE)
autoplot(fcst_arima, include = 120)
print(summary(fcst_arima))
```

One Step Ahead Sarima

NOTE doesn't run

```{r}
one_step_ahead_sarima = matrix(ncol = 2, nrow = 60)
for (i in 1:60){
  
  training_observed = window(data, start = c(1985,11), end = c(2016,(11+i)), frequency = 12)
  
  forecasted.sarima = sarima.for(training_observed,n.ahead=1,p=1,d=1,q=1,P=2,D=1,Q=1,S=12)
  
  demandforecast = forecasted.sarima$pred
  observed = validation[[i]]
  
  one_step_ahead_sarima[i,1]= observed
  one_step_ahead_sarima[i,2]= demandforecast
}
```


Capturing more than one year seasonality (period2=72)

NOTE the time element of the model changes 

```{r}
dshw_model_2 = dshw(sst_honolii_1985_2021_month_ts, period1=36, period2 =144, h = 420)

dshw_model_2


fcst_dshw <- forecast(dshw_model_2, h = 420)
autoplot(fcst_dshw, include = 420)
print(summary(fcst_dshw))

checkresiduals(fcst)
```

TBATS
