---
title: "Honolii_SST_Analysis_Forecasts"
output:
  pdf_document: default
  html_document: default
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

options(scipen = 999)
library(tidyverse)
library(ggplot2)
library(here)
library(lubridate)
library(forecast)
library(MLmetrics)
library(tsibble)
library(astsa)
library(janitor)
library(feasts)
library(gt)
library(modelr)
library(astsa)
library(tinytex)
```

Research Question - Has there been an upward trend in sea surface temperature in the 5km resolution grid box off of the Honolii coast? Can we use sst data, NOAA's coral bleaching model, and forecasting tools to:
  1. Analyze past coral bleaching events off the coast of Honolii, Hawaii for both heat tolerant and non heat tolerant corals
  2. Predict future coral bleaching events off the coast of Honolii, Hawaii for both heat tolerant and non heat tolerant corals

This exploratory analysis examines Sea Surface Temperature data from the NOAA Coral Reef Watch CoralTemp dataset. CoralTemp is a global sea surface temperature data product used primarily for coral bleaching monitoring. The dataset contains several data products based on 5 geostationary and 3 polar-orbiting satellites. The spatial resolution of the data is 5km with complete daily spatial coverage of the ocean from 1985-04-01 to present. The 5km grid for this analysis encompasses the marine ecosystem off of Honolii, Hawaii. 

First, we downloaded sea surface temperature data; calculated monthly means; and ran a linear regression model using ordinary least squares (OLS) to estimate the linear relationship between time and monthly mean sea surface temperature. 

```{r, results='hide'}
sst_honolii_1985_2021 <- read.csv(here("data", "sst_honolii_1985_2021.csv" ))
sst_honolii_1985_2021
```


```{r, results='hide'}
sst_honolii_1985_2021 <- sst_honolii_1985_2021 %>% 
  select(time, CRW_SST) %>% 
  mutate(time = as.Date(time))
sst_honolii_1985_2021
```


```{r, results='hide'}
sst_honolii_1985_2021_month <- sst_honolii_1985_2021 %>% 
  mutate(time = floor_date(time, "month")) %>% 
  group_by(time) %>% 
  summarise(mean_sst = mean(CRW_SST))

sst_honolii_1985_2021_month
```

```{r, results='hide'}
sst_honolii_1985_2021_month_plot <- ggplot(data = sst_honolii_1985_2021_month, aes(x = time, y = mean_sst)) +
  geom_line()

sst_honolii_1985_2021_month_plot
```

```{r}
sst_honolii_1985_2021_month_regression <- sst_honolii_1985_2021_month_plot +
  geom_smooth(method="lm", col="black")

sst_honolii_1985_2021_month_regression
```
```{r}
lm(mean_sst ~ time, data = sst_honolii_1985_2021_month) %>% 
  tidy %>% 
  gt()
```

We can see here that there is a positive correlation between time and mean sst. The coefficient on time tells us that for every one unit increase in time, the mean sst increases by 0.00002586144 degrees Celsius. The p value of 0.04 tells us this is statistically significant. 

Now we can check the residuals by plotting a histogram.
```{r}
ols <- lm(mean_sst ~ time, data = sst_honolii_1985_2021_month)
ols
summary(ols)

predictions <- sst_honolii_1985_2021_month %>% 
  add_predictions(ols) %>%
  mutate(residuals = mean_sst-pred)

# histogram
ggplot(data=predictions) + geom_histogram(aes(residuals), bins=25)

# mean
mean(predictions$residuals)
#1.495545e-14

# variance in residuals against mean_sst
ggplot(predictions) + geom_point(aes(x=mean_sst, y=residuals)) 

```

The residuals seem to be normally distributed.


```{r, results='hide'}
sst_honolii_1985_2021_month_ts <- ts(sst_honolii_1985_2021_month$mean_sst, start = c(1985, 11), frequency = 12)

sst_honolii_1985_2021_month_ts
```


```{r, results='hide'}
sst_honolii_1985_2021_month_tsibble <- sst_honolii_1985_2021_month_ts %>% as_tsibble() 
sst_honolii_1985_2021_month_tsibble
```

Next, we can run a classical decomposition model to analyze seasonality and trend

```{r}
sst_honolii_1985_2021_month_tsibble %>%
  model(
    classical_decomposition(value, type = "additive")
  ) %>%
  components() %>%
  autoplot() +
  labs(title = "Classical additive decomposition")
```
 Here, we can tell that seasonality plays a more important role in overall sst variation compared to trend. 

This marine ecosystem off the Hilo coast of the Big Island is known to house to coral that is showing resilliency toward warming sea surface temperatures caused by climate change. According to NOAA, corals experience stress if water reaches 1°C warmer than the highest expected annual temperature (Glynn and D'Croz, 1990). Thus, the bleaching threshold is defined as 1°C warmer than the maximum monthly mean temperature. To calculate the bleaching threshold, we first calculate the maximum monthly mean (mean sst of the month with the highest sst, for Hawaii this is September). 

```{r}
sst_honolii_1985_2021_month_tsibble %>% 
  gg_subseries(value)
```
\
From this visualization we can see the maximum monthly mean (depicted by the blue lines) for September is at 26.6 degrees Celsius. Therefore, we can calculate the bleaching threshold by adding 1 degree Celsius to the maximum monthly mean which gives us 27.6 degrees Celsius. 

Now, we can visualize any bleaching events over the last 35 years using daily sst. We can also do more! This marine ecosystem off the Hilo coast of the Big Island is known to house to coral that is showing resiliency toward warming sea surface temperatures caused by climate change. Coral can be heat tolerant for a variety of reasons. This analysis explores the hypothesis that the coral of Honolii reef have the ability to adapt to heat by hosting a symbiont algae that increases their bleaching threshold by approximately 1.5 degree Celsius compared to non heat tolerant coral. Therefore, our heat tolerant coral bleaching threshold is 29.1 degrees Celsius.  
```{r, results='hide'}
sst_honolii_1985_2021_plot <- ggplot(data = sst_honolii_1985_2021, aes(x = time, y = CRW_SST)) +
  geom_line()
sst_honolii_1985_2021_plot
```


```{r}
bleaching_threshold <- sst_honolii_1985_2021_plot +
  geom_hline(yintercept = 27.6, color = "red") + 
  geom_hline(yintercept = 29.1, color = "blue" )

bleaching_threshold
```

From this visualization, we can see that daily sst temperatures in the 5km grid box have surpassed the normal bleaching threshold in 4 warming events over the last 36 years, 3 of which occurred in the last 8 years. We can also see that heat tolerant coral, due to the symbiotic algae switch defined above, avoided all the bleaching events. 